A planetary collision afterglow and transit of the resultant debris cloud

### 2023-10-12 src/tex/ms.tex
[showyourwork] overleaf sync

%%==================================%%
%% sample for unstructured abstract %%
%%==================================%%

\abstract{
Planets grow in rotating disks of dust and gas around forming stars, some of which can subsequently collide in giant impacts after the gas component is removed from the disk \cite{Williams11,Wyatt15,Hughes18}.
%
Monitoring programs with the warm Spitzer mission have recorded significant and rapid changes in mid-infrared output for several stars, interpreted as variations in the surface area of warm dusty material ejected by planetary-scale collisions and heated by the central star: e.g., NGC 2354–ID8 \cite{2014Sci...345.1032M,Su19}, HD 166191 \cite{Su22} and V844 Persei \cite{2021ApJ...918...71R}.
%
%However, observational constraints on the aftermath and frequency of such impacts are difficult to obtain.
%
Here we report combined observations of the young ($\sim$~300~Myr), solar-like star \asas{}: an infrared brightening consistent with a blackbody temperature of 1000 K and a luminosity of 4\% of that of the star lasting for about 1000\,days, partially overlapping in time with a complex and deep wavelength-dependent optical eclipse that lasted for about 500 days.
%
The optical eclipse started 2.5\,years after the infrared brightening, implying an orbital period of at least that duration.
%
These observations are consistent with a collision between two exoplanets of several to tens of Earth masses at 2-16\,au from the central star. 
%
Such an impact produces a hot, highly-extended post-impact remnant with sufficient luminosity to explain the infrared observations.
%
Transit of the impact debris, sheared by orbital motion into a long cloud, causes the subsequent complex eclipse of the host star.
}

\keywords{exoplanets, planet formation, debris disks}
%TC:endignore

@@

% 262 words

@@

\begin{figure}
\begin{centering}
\includegraphics[width=\textwidth]{figures/all_phot_nature_delta_flux.pdf}
%\includegraphics[width=1.0\textwidth]{figures/master_lightcurve_nature_simple.pdf}
\caption{\textbf{Optical and infrared photometry of ASASSN-21qj.}
%
{\bf a}, Normalized optical photometry from \gls{asassn} in the $V$-band and the $g'$ band.
%
{\bf b},Fractional flux increase in brightness of ASASSN-21qj in both the $W1$ and $W2$ bands, for which a value of 1.0 represents the stellar contribution alone.
%
{\bf c},Calculated NEOWISE color temperature estimated from the photometry of the two bands.
%
The color temperature is plotted as zero when there is no infrared excess and is consistent with a temperature of 1000\,K while the excess is present.
%
Error bars are shown at $1\sigma$ confidence.
}
\label{fig:wisephot}
\script{plot_all_phot_nature_delta_flux.py}
%\script{plot_master_lightcurve_nature_simple.py}
\end{centering}
\end{figure}

% █ █▀█   █▀▄ █░█ █▀ ▀█▀   █░░ █▀█ █▀▀ ▄▀█ ▀█▀ █ █▀█ █▄░█
% █ █▀▄   █▄▀ █▄█ ▄█ ░█░   █▄▄ █▄█ █▄▄ █▀█ ░█░ █ █▄█ █░▀█

@@

% 205 words

@@

% 200 words

@@

% 284 words

@@

\begin{figure}[h!]
    \centering
\includegraphics[width=0.8\textwidth]{figures/sed.pdf}
    \caption{\textbf{Spectrum of \asas{}~components.} The red symbols show the optical and pre-brightnening \gls{wise} infrared photometry and the blue symbols show the post-brightnening \gls{wise} and \gls{alma} fluxes. Stellar and 1000\,K components consistent with the pre- and post-brightening fluxes are shown. The dashed line shows an estimated cool component spectrum for \SI{0.1}{\micro\metre}-sized grains associated with the transiting dust cloud. Downward triangles are upper limits.
    %
    Error bars are shown at $1\sigma$ confidence.
}
    \label{fig:sed}
   \script{fvsr.py}
\end{figure}
%

@@

Figure \ref{fig:eclipse_overview} shows the optical light curves. The start of the optical eclipse is seen around MJD 59350 in the $g'$ band observations, and returns to pre-eclipse levels by MJD 59850, giving a total eclipse duration of approximately 500 days. The eclipse depth varies as a function of wavelength, which is discussed below.
%
The normalisation of the light curves for the ASAS-SN, ATLAS and ALLWISE photometry was by calculating the out of transit flux before MJD 58700.
%
The LCOGT and AAVSO photometry was determined by aligning the photometry of the eclipse with the ASAS-SN and ATLAS photometry on a per band basis.
%
By treating the linear changes in flux as due to the edges of large dust clouds crossing the disk of the star, and the largest amount of absorption within any of these segments as an estimate for the absorption of the cloud \citep[see equations 4.2 and 4.3 in ][]{Kennedy17}, we can determine a robust lower limit to the transverse velocity of 7.5 km\,s$^{-1}$ for the material moving in front of the star. The method is illustrated in Extended Data Figure~\ref{fig:gradientconvert} and the results over the full optical transit are shown in the lower panel of Figure~\ref{fig:eclipse_overview}.
%
We convert the magnitude $M(t)$ at time $t$ to a normalised flux $f(t)=10^{(M(t)-M_0/-2.5)}$ where $M_0$ is the mean magnitude outside of the eclipse.
%
We visually determine turning points in the linearly increasing or decreasing photometric flux, fit straight lines to the selected points, and determine the flux gradient in units of day$^{-1}$.
%
A lower bound can be derived for the transverse velocity of the dust, $v$, by measuring the gradient of the light curve and determining what velocity a sharp edged and completely opaque occulter moving across the disk of the star would be needed to make the same gradient.
%
If the dust is on a circular orbit, it therefore has to be within 16\,au (equivalent to an orbital period of 63 years) around the star. The temperature of sub-micron dust grains at this distance would be approximately 200\,K.

%\subsection*{Dust properties from the colors in the optical} MAX 40 characters
\subsubsection*{Dust properties from optical colors}

@@

As is clear from Figure \ref{fig:eclipse_overview} the photometry shows deeper absorption at bluer wavelengths compared to redder wavelengths.
%
This wavelength dependent absorption is typical of extinction due to particles with a characteristic size similar to or smaller than that of the observed wavelengths.
%This difference is an indication that the extinction is caused by particles equal to or smaller than the wavelengths of light.
%
The differences are quantified in Extended Data Figure \ref{fig:blueing}, which shows the AAVSO $BVI$ photometric colours as a function of $V$ magnitude.
%
For both $B-V$ and $V-I$ the colour becomes redder as the star dims, and the reddening is here quantified by the total to selective extinction ratio $A_\lambda/A_V$ (dashed lines).
%
The values are similar to those seen for interstellar extinction \cite{1989ApJ...345..245C}, indicating that the dimming is caused by sub-micron sized dust \cite{1977ApJ...217..425M,2001ApJ...548..296W}.
%
A further constraint from the colours relates to scattering; in protoplanetary disk systems that undergo dimming (e.g. UX~Ori types) the colour initially reddens with dimming, but moves back towards the stellar colour when the dimming is more than one or two magnitudes \cite[e.g.][]{1994AJ....108.1906H}.
%
This ``blueing'' is typically interpreted as the relative increase in dust-scattered starlight from the disk or envelope as the star itself fades \cite{1988SvAL...14...27G}.
%
This behaviour is also seen in Extended Data Figure \ref{fig:blueing}, where $B-V$ shows significant blueing, while $V-I$ does not.
%
The solid lines show an extinction model where an underlying scattered light component with the same colour as the star has been added; as the star dims it reddens, but will eventually return to the stellar colour.
%
This happens more quickly for $B-V$ because nearly all of the stellar flux in $B$ is blocked, and is less pronounced for $V-I$ because the star is significantly less dimmed in $I$. The fraction of scattered light is relatively large at 7.5\%, implying a significant complex of small dust around the star by the time the deepest parts of the optical transit occur. This high fraction suggests that the impact occurred a significant fraction of an orbit before the optical transit, thus allowing the dust complex time to spread around the star.

@@

\begin{figure}
    \centering
\includegraphics[width=1.0\textwidth]{figures/final_v2_less0d029.jpg}
    \caption{\textbf{Simulations of the formation of a post-impact body.}
    %
    Giant impacts between super-Earths and mini-Neptunes can produce post-impact bodies hundreds of Earth-radii across, comparable with that required to produce the observed infrared flux. With the exception of the lower right panel, particles are colored by their material (forsterite, water, or a H$_2$-He mixture moving outwards in the initial bodies) and whether they came from the impactor or target (see top left panel). The final two panels show just the mass bound to the primary remnant which has a mass of 48.4~$M_{\rm Earth}$. In the final panel, particles that are at the minimum density imposed by the code are colored in green.}
    \label{fig:SPH}
 %   \script{fvsr.py}
\end{figure}

@@

%Acknowledgments are not compulsory. Where included they should be brief. Grant or contribution numbers may be acknowledged.
G.M.K is supported by the Royal Society as a Royal Society University Research Fellow.
%
S. J. L. acknowledges funding from the UK Natural Environment Research Council (grant NE/V014129/1).
%
L.C. acknowledges funding from the European Union H2020-MSCA-ITN-2019 under Grant Agreement no. 860470 (CHAMELEON)
%
J. D. acknowledges funding support from the Chinese Scholarship Council (No. 202008610218). Giant impact simulations were carried out using the Isambard 2 UK National Tier-2 HPC Service (http://gw4.ac.uk/isambard/) operated by GW4 and the UK Met Office, and funded by EPSRC (EP/T022078/1).
%
We thank Krzysztof Stanek and the work of the ASAS-SN team with their survey and for providing public access to the database.
%
Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration (80NM0018D0004).

@@

\bmhead{Code availability} 
All the code for the analysis and the generation of all the figures  are available in a showyourwork \citep{Luger2021} reproducible framework available as a git repository on github. \url{https://github.com/mkenworthy/ASASSN-21qj-debris/}. The source code and documentation for the SWIFT open-source simulation code is available from \url{www.swiftsim.com}.

%\bmhead{Supplementary information}
%Supplementary Information is available for this paper.

>>>>

\abstract{
Planets grow in rotating disks of dust and gas around forming stars, some of which can subsequently collide in giant impacts after the gas component is removed from the disk \cite{Williams11,Wyatt15,Hughes18}.
%
Monitoring programs with the warm Spitzer mission have recorded significant and rapid changes in mid-infrared output for several stars, interpreted as variations in the surface area of warm dusty material ejected by planetary-scale collisions and heated by the central star: e.g., NGC 2354–ID8 \cite{2014Sci...345.1032M,Su19}, HD 166191 \cite{Su22} and V844 Persei \cite{2021ApJ...918...71R}.
%
Here we report combined observations of the young ($\sim$~300~Myr), solar-like star \asas{}: an infrared brightening consistent with a blackbody temperature of 1000 K and a luminosity of 4\% of that of the star lasting for about 1000\,days, partially overlapping in time with a complex and deep wavelength-dependent optical eclipse that lasted for about 500 days.
%
The optical eclipse started 2.5\,years after the infrared brightening, implying an orbital period of at least that duration.
%
These observations are consistent with a collision between two exoplanets of several to tens of Earth masses at 2-16\,au from the central star. 
%
Such an impact produces a hot, highly-extended post-impact remnant with sufficient luminosity to explain the infrared observations.
%
Transit of the impact debris, sheared by orbital motion into a long cloud, causes the subsequent complex eclipse of the host star.
}

\keywords{exoplanets, planet formation, debris disks}

@@

@@

\begin{figure}
\begin{centering}
\includegraphics[width=\textwidth]{figures/all_phot_nature_delta_flux.pdf}
\caption{\textbf{Optical and infrared photometry of ASASSN-21qj.}
%
{\bf a}, Normalized optical photometry from \gls{asassn} in the $V$-band and the $g'$ band.
%
{\bf b},Fractional flux increase in brightness of ASASSN-21qj in both the $W1$ and $W2$ bands, for which a value of 1.0 represents the stellar contribution alone.
%
{\bf c},Calculated NEOWISE color temperature estimated from the photometry of the two bands.
%
The color temperature is plotted as zero when there is no infrared excess and is consistent with a temperature of 1000\,K while the excess is present.
%
Error bars are shown at $1\sigma$ confidence.
}
\label{fig:wisephot}
\script{plot_all_phot_nature_delta_flux.py}
\end{centering}
\end{figure}

@@

@@

@@

@@

\begin{figure}[h!]
    \centering
\includegraphics[width=0.8\textwidth]{figures/sed.pdf}
    \caption{\textbf{Spectrum of \asas{}~components.}
    %
    The red symbols show the optical and pre-brightnening \gls{wise} infrared photometry and the blue symbols show the post-brightnening \gls{wise} and \gls{alma} fluxes.
    %
    Stellar and 1000\,K components consistent with the pre- and post-brightening fluxes are shown.
    %
    The dashed line shows an estimated cool component spectrum for \SI{0.1}{\micro\metre}-sized grains associated with the transiting dust cloud.
    %
    Downward triangles are upper limits.
    %
    Error bars are shown at $1\sigma$ confidence.
}
    \label{fig:sed}
   \script{fvsr.py}
\end{figure}
%

@@

Figure \ref{fig:eclipse_overview} shows the optical light curves.
%
The start of the optical eclipse is seen around MJD 59350 in the $g'$ band observations, and returns to pre-eclipse levels by MJD 59850, giving a total eclipse duration of approximately 500 days.
%
The eclipse depth varies as a function of wavelength, which is discussed below.
%
The normalisation of the light curves for the ASAS-SN, ATLAS and ALLWISE photometry was by calculating the out of transit flux before MJD 58700.
%
The LCOGT and AAVSO photometry was determined by aligning the photometry of the eclipse with the ASAS-SN and ATLAS photometry on a per band basis.
%
By treating the linear changes in flux as due to the edges of large dust clouds crossing the disk of the star, and the largest amount of absorption within any of these segments as an estimate for the absorption of the cloud \citep[see equations 4.2 and 4.3 in ][]{Kennedy17}, we can determine a robust lower limit to the transverse velocity of 7.5 km\,s$^{-1}$ for the material moving in front of the star.
%
The method is illustrated in Extended Data Figure~\ref{fig:gradientconvert} and the results over the full optical transit are shown in the lower panel of Figure~\ref{fig:eclipse_overview}.
%
We convert the magnitude $M(t)$ at time $t$ to a normalised flux $f(t)=10^{(M(t)-M_0/-2.5)}$ where $M_0$ is the mean magnitude outside of the eclipse.
%
We visually determine turning points in the linearly increasing or decreasing photometric flux, fit straight lines to the selected points, and determine the flux gradient in units of day$^{-1}$.
%
A lower bound can be derived for the transverse velocity of the dust, $v$, by measuring the gradient of the light curve and determining what velocity a sharp edged and completely opaque occulter moving across the disk of the star would be needed to make the same gradient.
%
If the dust is on a circular orbit, it therefore has to be within 16\,au (equivalent to an orbital period of 63 years) around the star. The temperature of sub-micron dust grains at this distance would be approximately 200\,K.

\subsubsection*{Dust properties from optical colors}

@@

As is clear from Figure \ref{fig:eclipse_overview} the photometry shows deeper absorption at bluer wavelengths compared to redder wavelengths.
%
This wavelength dependent absorption is typical of extinction due to particles with a characteristic size similar to or smaller than that of the observed wavelengths.
%
The differences are quantified in Extended Data Figure \ref{fig:blueing}, which shows the AAVSO $BVI$ photometric colours as a function of $V$ magnitude.
%
For both $B-V$ and $V-I$ the colour becomes redder as the star dims, and the reddening is here quantified by the total to selective extinction ratio $A_\lambda/A_V$ (dashed lines).
%
The values are similar to those seen for interstellar extinction \cite{1989ApJ...345..245C}, indicating that the dimming is caused by sub-micron sized dust \cite{1977ApJ...217..425M,2001ApJ...548..296W}.
%
A further constraint from the colours relates to scattering; in protoplanetary disk systems that undergo dimming (e.g. UX~Ori types) the colour initially reddens with dimming, but moves back towards the stellar colour when the dimming is more than one or two magnitudes \cite[e.g.][]{1994AJ....108.1906H}.
%
This ``blueing'' is typically interpreted as the relative increase in dust-scattered starlight from the disk or envelope as the star itself fades \cite{1988SvAL...14...27G}.
%
This behaviour is also seen in Extended Data Figure \ref{fig:blueing}, where $B-V$ shows significant blueing, while $V-I$ does not.
%
The solid lines show an extinction model where an underlying scattered light component with the same colour as the star has been added; as the star dims it reddens, but will eventually return to the stellar colour.
%
This happens more quickly for $B-V$ because nearly all of the stellar flux in $B$ is blocked, and is less pronounced for $V-I$ because the star is significantly less dimmed in $I$.
%
The fraction of scattered light is relatively large at 7.5\%, implying a significant complex of small dust around the star by the time the deepest parts of the optical transit occur.
%
This high fraction suggests that the impact occurred a significant fraction of an orbit before the optical transit, thus allowing the dust complex time to spread around the star.

@@

\begin{figure}
    \centering
\includegraphics[width=1.0\textwidth]{figures/final_v2_less0d029.jpg}
    \caption{\textbf{Simulations of the formation of a post-impact body.}
    %
    Giant impacts between super-Earths and mini-Neptunes can produce post-impact bodies hundreds of Earth-radii across, comparable with that required to produce the observed infrared flux.
    %
    With the exception of the lower right panel, particles are colored by their material (forsterite, water, or a H$_2$-He mixture moving outwards in the initial bodies) and whether they came from the impactor or target (see top left panel).
    %
    The final two panels show just the mass bound to the primary remnant which has a mass of 48.4~$M_{\rm Earth}$.
    %
    In the final panel, particles that are at the minimum density imposed by the code are colored in green.}
    \label{fig:SPH}
\end{figure}

@@

G.M.K is supported by the Royal Society as a Royal Society University Research Fellow.
%
S. J. L. acknowledges funding from the UK Natural Environment Research Council (grant NE/V014129/1).
%
L.C. acknowledges funding from the European Union H2020-MSCA-ITN-2019 under Grant Agreement no. 860470 (CHAMELEON)
%
J. D. acknowledges funding support from the Chinese Scholarship Council (No. 202008610218).
%
Giant impact simulations were carried out using the Isambard 2 UK National Tier-2 HPC Service (http://gw4.ac.uk/isambard/) operated by GW4 and the UK Met Office, and funded by EPSRC (EP/T022078/1).
%
We thank Krzysztof Stanek and the work of the ASAS-SN team with their survey and for providing public access to the database.
%
Part of this research was carried out at the Jet Propulsion Laboratory, California Institute of Technology, under a contract with the National Aeronautics and Space Administration (80NM0018D0004).

@@

\bmhead{Code availability} 
All the code for the analysis and the generation of all the figures  are available in a showyourwork \citep{Luger2021} reproducible framework available as a git repository on github: \url{https://github.com/mkenworthy/ASASSN-21qj-debris/}.
%
The source code and documentation for the SWIFT open-source simulation code is available from \url{www.swiftsim.com}.
